---
title: "RNPC3 minor intron splicing analysis"
author: "Xueyi Dong"
date: '2022-05-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

### Samples information
A549 cell line samples provided by Joan Heath, sequenced using nanopore PCR cDNA.
barcode01-04: controls (non-targeted), barcode05-08: targeted with rnpc3 siRNA 18.

### Pre-processing
Mapped to human CHM13v2 reference genome using minimap2 version 2.17.
Novel isoforms identified and counts generated using bambu version 1.0.3.
The gtf was then compared to CHM13v2 annotation using gffcompare.

### Aims and plan
We want to explore human minor introns and splicing events using long read sequencing.
First, we will check the novel isoforms detected. We will especially focus on whether there is "intron retention" class novel isoforms from the gene list provided by Joan. Next, we will perform a differential transcript expression analysis and a differential transcript usage analysis.

## Prepare and load the data

```{r setup}
library(edgeR)
```

```{r load_data}
counts <- read.delim("../bambu/out/counts_transcript.txt")
rownames(counts) <- counts$TXNAME
tmap <- read.delim("../bambu/out/bambu_comp.extended_annotations.gtf.tmap")
mi_gene <- xlsx::read.xlsx("../HomoSapiens_IntronInfo.xlsx", sheetIndex = 1)
```

## minor intron gene analysis

Check which minor intron genes are not available in our dataset.
```{r}
mi_gene$Gene.Name[!(mi_gene$Gene.Name %in% tmap$ref_gene_id)]
```

We extract transcript class "m" and "n" (intron retentions) from gffcompare output.
```{r}
tmap.m <- tmap[tmap$class_code == "m", ]
tmap.n <- tmap[tmap$class_code == "n", ]
```

Check how many of "m" and "n" are in the minor intron genes list. We use "ref_gene_id" (annotated gene name) to identify the gene that the isoforms belong to.
```{r}
table(tmap.m$ref_gene_id %in% mi_gene$Gene.Name)
table(tmap.n$ref_gene_id %in% mi_gene$Gene.Name)
```

Save the lists of "m" and "n" minor intron isoforms.
```{r}
write.csv(tmap.m[tmap.m$ref_gene_id %in% mi_gene$Gene.Name, ],
          "results/miGenesClassMTx.csv")
write.csv(tmap.n[tmap.n$ref_gene_id %in% mi_gene$Gene.Name, ],
          "results/miGenesClassNTx.csv")
```

The transcript ids are in the "qry_id" column of the lists.

[class m](results/miGenesClassMTx.csv)
[class n](results/miGenesClassNTx.csv)

## Differential expression analysis

We use edgeR quasi-likelihood method for DE analysis.

```{r}

x <- DGEList(counts = counts[, c(-1, -2)])

```

